Safe Memory Regions for Big Data Processing

### 2016-10-19 paper/related.tex
related work

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically enforce the safety of
their region-based memory management scheme in ML. Following their
seminal work, static type systems for safe region-based memory
management have been extensively studied in the context of various
languages and problem settings~\cite{cyclone02,cyclone04,yates99,MIT03,DPJ09,HMN01,WW01,rust,gpu14}.
Our work differs from the
existing proposals in one or more of the following respects.

@@

\item We adopt a two-pronged approach to memory safety that relies on a
combination of a simple static type discipline and lightweight runtime
checks. In particular, our approach circumvents the need for restrictive
static mechanisms (e.g., linear types and unique pointers) or
expensive runtime mechanisms (e.g., garbage collection and reference
counting) in order to guarantee safety.

@@

Tofte and Talpin's approach~\cite{tofte97} uses compiler-managed lexically
scoped (stack) regions (as a replacement for GC).
% Similar to their region inference algorithm, our region type inference
% algorithm can deal with region-polymorphic recursion by computing
% fixed points for recursive constraints. 
Our type inference is analogous to their region inference in some respects,
while differing in others. While their inference algorithm only generates equality constraints, which
can be solved via unification, our type inference algorithm also
generates partial order outlives constraints, which are required to
capture relationships between lifetimes of transferable regions
and stack regions. Consequently, our constraint solving algorithm is
more sophisticated, and is capable of inferring unknown outlives
constraints over region arguments of polymorphic recursive functions.

Walker and Watkins~\cite{WW01} extend lambda calculus with first-class regions
with dynamic lifetimes, and imposes linear typing to control accesses to regions.
Our open/close lexical block for transferable regions traces its origins to the \C{let!}
expression in~\cite{WW01} and~\cite{wadler90}, which safely relaxes
linear typing restrictions, allowing variables to be temporarily
aliased.
We don't use linear typing (for references to regions), thus admit unrestricted aliasing,
but use lightweight runtime checks for safety.
% Non-linear typing keeps our type system simple, and makes it
% possible to infer types, thus eliminating the annotation burden.
Moreover, ~\cite{WW01}'s linear type system is insufficient to enforce
the invariants needed to ensure safety under region transfers, such as
the absence of references that escape a transferable region.

Cyclone~\cite{cyclone02} equips C with programmer-managed stack
regions, and a typing discipline that statically guarantees the safety
of all pointer dereferences. Later
proposals~\cite{cyclone04,cycloneSCP} extends Cyclone with dynamic
regions. \name differs from Cyclone fundamentally because of its
non-intrusiveness design principle, which requires its safety
mechanisms to not intrude on the the programming practices of C\#.
\name programmers, for example, shouldn’t be forced to abandon
iterators in favor of for-loops, annotate region types, or rewrite
C\#'s standard libraries to use in \name. Cyclone requires C
programmers to use new language constructs and abandon some standard
programming idioms in the interest of preserving safety. For instance,
Cyclone programmers are required to write region types for functions;
the type inference is only intraprocedural. Ensuring safety in
presence of dynamic regions requires using either unique pointers or
reference-counted objects.  Both approaches are intrusive. For
example, unique pointers constrain, or in some cases forbid, the use
of the familiar iterator pattern, which requires creation of aliases
to objects in a collection. Some standard library functions, for
example, those that use caching, may need to be rewritten.  Moreover,
even with unique pointers, safety cannot be guaranteed statically;
checks against \C{NULL} are needed at run-time to enforce safety. For
ref-counted objects, Cyclone requires programmers to use special
functions (\C{alias\_refptr} and \C{drop\_refptr}) to create and
destroy aliases.  Reference count is affected only by these functions.
An alias going out of scope, for instance, does not decrement the
ref-count. The requirement to use additional constructs
to manage aliases makes reference counting more-or-less as intrusive
as unique pointers.

@@

Our region type system can also be thought of as a specialized
ownership type system~\cite{OwnershipSurvey}, where each region is the
owner of all objects allocated in the region.
%
An ownership type system for safe region-based memory management in
real-time Java has been proposed by~\cite{MIT03}.  
% Like us, they too assume a source language with programmer-managed
% memory regions, and focus on proving safety of programs written in
% that language.  
Their language permits only lexically-scoped (stack) regions
(in the context of shared-memory concurrency). In contrast, we permit
regions with dynamically determined lifetimes (in the context of message-passing concurrency).
We borrow outlives relation from their formal
development, and our type system bears some similarities to theirs.
However, our language also admits parametric polymorphism (generics)
and higher-order functions.
Moreover, we establish type safety and transfer safety results that
formalize the guarantees provided by our system. While~\cite{MIT03}'s
language is explicitly typed, our language comes equipped with full
type inference.
% Although there is some support for local
% type inference, region types for methods and classes need to be
% written explicitly. In contrast, our region type inference that
% eliminates any such need.
% Our type inference algorithm is also novel compared to existing type
% inference approaches for ownership types, which are based on, e.g.,
% context-insensitive pointer analysis or boolean satisfiability.
% E.g., papers 55 and 79.

~\cite{HMN01} proposes a flow-sensitive region-based memory management
for first-order programs that proposes to overcome some of the
drawbacks of~\cite{tofte97} by generalizing~\cite{tofte97}'s approach
to regions with dynamic lifetimes. However, dynamic regions are still
not first-class values of the language, and reference counting is
nonetheless needed to ensure memory safety.

The idea of using region-based memory management to facilitate the
safe transfer of rich data structures between computational nodes has
been previously explored by~\cite{gpu14} in the context of Scheme
language. However, their setting only includes lexically-scoped
regions for which Tofte and Talpin-style analysis~\cite{tofte97}
suffices. In contrast, our language provides first-class support for
transferable regions with dynamic lifetimes. We require this
generality in order to support streaming query operators, such as the
one shown in Fig.~\ref{fig:motivating-eg}.

>>>>

% Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
% concept of a region type system to statically enforce the safety of
% their region-based memory management scheme in ML.
Following Tofte and Talpin's seminal work in~\cite{tofte93,tofte94,tofte97},
static type systems for safe region-based memory
management have been extensively studied in the context of various
languages and problem settings~\cite{cyclone02,cyclone04,yates99,MIT03,DPJ09,HMN01,WW01,rust,gpu14}.
Our work differs from the
existing proposals in one or more of the following respects.

@@

\item We exploit a combination of a simple static type discipline and lightweight
  runtime checks to ensure memory safety.
  % adopt a two-pronged approach to memory safety that relies on a
In particular, our approach circumvents the need for restrictive
static mechanisms (e.g., linear types and unique pointers) or
expensive runtime mechanisms (e.g., garbage collection and reference
counting) in order to guarantee safety.

@@

Tofte and Talpin's approach~\cite{tofte97} uses compiler-managed lexically
scoped (stack) regions (as a replacement for GC).
Our type inference is analogous to their's in some respects,
while differing in others.
Their inference algorithm only generates equality constraints, solvable via unification.
Our type inference algorithm generates partial order outlives constraints.
% which are required to capture relationships between lifetimes of transferable regions and stack regions.
Consequently, our constraint solving algorithm is
more sophisticated, and is capable of inferring unknown outlives
constraints over region arguments of polymorphic recursive functions.

Walker and Watkins~\cite{WW01} extend lambda calculus with first-class regions
with dynamic lifetimes, and impose linear typing to control accesses to regions.
Our open/close lexical block for transferable regions traces its origins to the \C{let!}
expression in~\cite{WW01} and~\cite{wadler90}, which safely relaxes
linear typing restrictions, allowing variables to be temporarily
aliased.
We don't use linear typing (for references to regions), thus admit unrestricted aliasing,
but use lightweight runtime checks for safety.
% Non-linear typing keeps our type system simple, and makes it
% possible to infer types, thus eliminating the annotation burden.
Moreover, ~\cite{WW01}'s linear type system is insufficient to enforce
the invariants needed to ensure safety under region transfers, such as
the absence of references that escape a transferable region.

Cyclone~\cite{cyclone02} equips C with programmer-managed stack
regions, and a typing discipline that statically guarantees the safety
of all pointer dereferences. Later
proposals~\cite{cyclone04,cycloneSCP} extends Cyclone with dynamic
regions. \name differs from Cyclone in its non-intrusiveness design principle,
which requires its safety
mechanisms to not intrude on the programming practices of C\#.
\name programmers, for example, shouldn't be forced to abandon
iterators in favor of for-loops, annotate region types, or rewrite
C\#'s standard libraries to use in \name. Cyclone requires C
programmers to use new language constructs and abandon some standard
programming idioms in the interest of preserving safety. For instance,
Cyclone programmers are required to write region types for functions;
the type inference is only intraprocedural. Ensuring safety in
presence of dynamic regions requires using either unique pointers or
reference-counted objects.  Both approaches are intrusive. For
example, unique pointers constrain, or in some cases forbid, the use
of the familiar iterator pattern, which requires creation of aliases
to objects in a collection. Some standard library functions, for
example, those that use caching, may need to be rewritten.  Moreover,
even with unique pointers, safety cannot be guaranteed statically;
checks against \C{NULL} are needed at run-time to enforce safety. For
ref-counted objects, Cyclone requires programmers to use special
functions (\C{alias\_refptr} and \C{drop\_refptr}) to create and
destroy aliases.  Reference count is affected only by these functions.
An alias going out of scope, for instance, does not decrement the
ref-count. The requirement to use additional constructs
to manage aliases makes reference counting more-or-less as intrusive
as unique pointers.

@@

Our region type system can also be thought of as a specialized
ownership type system~\cite{OwnershipSurvey}, where each region is the
owner of all objects allocated in the region.
%
An ownership type system for safe region-based memory management in
real-time Java has been proposed by~\cite{MIT03}.  
% Like us, they too assume a source language with programmer-managed
% memory regions, and focus on proving safety of programs written in
% that language.  
Their language permits only lexically-scoped (stack) regions
(in the context of shared-memory concurrency). In contrast, we permit
regions with dynamically determined lifetimes (in the context of message-passing concurrency).
% We borrow outlives relation from their formal
% development, and our type system bears some similarities to theirs.
Our language also admits generics and higher-order functions.
We also establish type safety and transfer safety results that
formalize the guarantees provided by our system. While~\cite{MIT03}'s
language is explicitly typed, our language comes equipped with full
type inference.

% Although there is some support for local
% type inference, region types for methods and classes need to be
% written explicitly. In contrast, our region type inference that
% eliminates any such need.
% Our type inference algorithm is also novel compared to existing type
% inference approaches for ownership types, which are based on, e.g.,
% context-insensitive pointer analysis or boolean satisfiability.
% E.g., papers 55 and 79.

~\cite{HMN01} proposes a flow-sensitive approach for first-order programs
% to overcome some of the drawbacks of~\cite{tofte97}
to generalize~\cite{tofte97}'s approach to dynamic regions.
However, dynamic regions are still not first-class values of the language,
and reference counting is needed to ensure memory safety.
%
\cite{gpu14} uses regions to safely transfer data between the CPU and GPU
in the context of Scheme. However, their setting only includes lexically-scoped
regions for which Tofte and Talpin-style analysis~\cite{tofte97}
suffices. In contrast, we provide first-class support for
transferable regions with dynamic lifetimes.
% We require this generality in order to support streaming query operators, such as the
% one shown in Fig.~\ref{fig:motivating-eg}.
